Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-04-09 manuscript.tex
next round of edits.

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig4Boxplots.pdf}
    \caption{Box-plot showing tillering index in our greenhouse population from day 20-60. White indicates individuals homozygous for the Hopscotch, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No Hopscotch) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles.} 
\label{Fig4Boxplots}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the Hopscotch allele segregating at a frequency of 0.69, \jri{does this violate HWE? didn't we choose numbers of each genotype here, so that HWE isn't appropriate? if so, probably should drop genotype frequency data here} with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

@@

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \citep{Clark2006, Studer2011}, \jri{did clark do phenotyping?} and \citet{Weber2007} have described its effects in partially inbred lines of teosinte \citep{Weber2007}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in individuals sampled from a natural population of teosinte. However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize. \lev{i took out the information about the TILS -- seemed the consensus was more than we wanted to do for this pub. If qPCR is working for tb1 it would be something easy to check...} It is certainly possible that even though previous data demonstrate an effect of the Hopscotch on tillering in maize \citep{Studer2011}, that the effect of the Hopscotch in teosinte is more complicated and may be more difficult to observe. Moreover, \emph{tb1} is a single gene in a complex pathway that affects branching and tillering traits, and perhaps in combination with alleles at other loci the phenotypic effects of the Hopscotch on tillering may not be consistent.

Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012, PAPERS}, and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

In summary, our findings demonstrate that the Hopscotch allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the Hopscotch allele is present in these populations due to recent introgression; however, it seems unlikely that the it would have drifted to high frequency in multiple populations, and there may be another explanation for the high frequency we observe in some of our populations. The Hopscotch does not appear to uniformly reduce tillering in teosinte as it does in maize, and the other loci involved in branching architecture of maize may play key roles in the regulation of tillering in teosinte. Furthermore, the high frequency with which we find the Hopscotch in a number of populations may suggest that the Hopscotch may play an additional ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjuction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? these genes need citations.}

>>>>

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig4Boxplots.pdf}
    \caption{Box-plot showing tillering index in our greenhouse population from day 20-60. White indicates individuals homozygous for the Hopscotch, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No Hopscotch) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles.} 
\label{Fig4Boxplots}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------
\jri{lots of white space in fig 4 and fig. s1 too.} 
We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the Hopscotch allele segregating at a frequency of 0.69, \jri{does this violate HWE? didn't we choose numbers of each genotype here, so that HWE isn't appropriate? if so, probably should drop genotype frequency data here} with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

@@

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \citep{Clark2006, Studer2011}, \jri{did clark do phenotyping?} and \citet{Weber2007} have described its effects in partially inbred lines of teosinte \citep{Weber2007} \jri{i don't think these were inbred.please doublecheck.}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in individuals sampled from a natural population of teosinte. \jri{isn't this what weber did?? for 70+ populations!?} However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize. It is certainly possible that even though previous data demonstrate an effect of the Hopscotch on tillering in maize \citep{Studer2011}, that the effect of the Hopscotch in teosinte is more complicated and may be more difficult to observe. Moreover, \emph{tb1} is a single gene in a complex pathway that affects branching and tillering traits, and perhaps in combination with alleles at other loci the phenotypic effects of the Hopscotch on tillering may not be consistent. \jri{thsi section still needs work. i think we have to do more here. weber shows an association between SNPs in tb1 and branch length. we need to discuss that!}

\jri{MBH todo} Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012, PAPERS}, \jri{yup, cite some PAPERS} and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

In summary, our findings demonstrate that the Hopscotch allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the Hopscotch allele is present in these populations due to recent introgression; however, it seems unlikely that  it would have drifted to high frequency in multiple populations and there may be another explanation for the high frequency we observe in some of our populations. The Hopscotch does not appear to have a strong effect reducing tillering in teosinte as it does in maize, and  other loci involved in branching architecture  may play also play roles in the regulation of tillering in teosinte. Finally, although we see no clear evidence of recent strong selection, the high frequency of the Hopscotch insertion in a number of populations continues to suggest to us that it plays an ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjunction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? }

\jri{please check format of supp figs and tables; some are running off the page.  check fig/table references, bibliography, etc. what does "rotation" mean in supp. table 3? it isn't mentioned in methods. please check that all the tables and figs (including supplement) are referenced in the text.}
